\documentclass[11pt,a4paper]{article}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{geometry}
\usepackage{booktabs}

\geometry{left=2.5cm, right=2.5cm, top=2.5cm, bottom=2.5cm}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}

\begin{document}

\title{Autonomous Multiclass TRM: Unified Spatial Kinematics and Isomorphic Moving Bottleneck Subclass Extension}
\author{Mingchen Yuan}
\date{}
\maketitle

\section{Introduction}
This paper presents a complete 3D traffic flow model that combines basic traffic laws with a special extension for moving bottlenecks. It accurately models how heavy trucks can trap faster passenger cars. By using physics-based equations and a step-by-step calculation method, this model solves the complex problem of localized traffic jams smoothly, without crashing the computer simulation or creating unrealistic vehicle movements.

\section{Universal State Variables and Isomorphic Subclasses}
Let the highway be divided into $X$ cells $x\in\{1, \dots, X\}$ of length $\Delta x$, and parallel lanes $l\in\{1, \dots, L\}$. Vehicles travel at discrete speed levels $i \in \{1, \dots, N\}$, representing actual speeds $v_1 < v_2 < \dots < v_N$. 

To comprehensively model the moving bottleneck effect (where heavy trucks trap faster cars), we divide the traffic into three distinct groups ($m$):
\begin{itemize}
    \item \textbf{Class A (Trucks):} Heavy vehicles with a free-flow speed limit of $v_A^{ff}$.
    \item \textbf{Class Bf (B-fast):} Freely flowing passenger cars.
    \item \textbf{Class Bs (B-slow):} Passenger cars strictly trapped behind trucks.
\end{itemize}

To prevent vehicles from unrealistically overlapping in the simulation, the road's physical capacity limit $\rho_{\max}$ is governed by the \textbf{Effective Spatial Occupancy Density}. This is calculated using a size weight $w^{(m)}$ (Passenger Car Equivalent) for each vehicle type:
\begin{equation}
\Omega_{x,l} = \sum_{m\in\{A,Bf,Bs\}}\sum_{i=1}^{N} w^{(m)} f_{i,x,l}^{(m)}
\end{equation}
where $w^{(Bf)} = w^{(Bs)} = w^{(B)}$. This ensures that a car's physical size remains exactly the same whether it is flowing freely or trapped.

\subsection{Strict Rules for Trapped Cars}
The highest speed index a truck can reach is $i_{thr} = \max\{i : v_i \le v_A^{ff}\}$. Both free cars ($Bf$) and trapped cars ($Bs$) exist across all speed levels $i \in \{1, \dots, N\}$. To mathematically simulate being "trapped," we apply two absolute rules exclusively to Class Bs:

\textbf{1. Acceleration Blockade:} A trapped car is strictly forbidden from accelerating past the truck's speed limit. However, normal deceleration is still allowed:
\begin{equation}
\lambda_{i\rightarrow i+1}^{(Bs)} \equiv 0, \quad \forall i \ge i_{thr}
\end{equation}

\textbf{2. Absolute Lateral Immobilization:} By physical definition, a car is only trapped if it cannot escape to another lane. Therefore, all lateral lane-changing movements ($\gamma$) for trapped cars are frozen until they are officially released:
\begin{equation}
\gamma_{l\rightarrow l', i}^{(Bs)} \equiv 0, \quad \forall l, l'
\end{equation}

\section{Baseline Internal Kinematics and Lateral Dynamics}
For normal, freely moving traffic (Class A and Bf, and for deceleration in Bs), speed changes only happen between adjacent speed levels ($|i-j|\le 1$). They cannot exceed the lowest or highest speed limits ($\lambda_{1\rightarrow 0} \equiv 0$, $\lambda_{N\rightarrow N+1} \equiv 0$). 

\subsection{Acceleration and Deceleration ($\lambda$)}
A vehicle's acceleration capability decreases as it approaches its maximum physical speed:
\begin{equation}
\lambda_{i\rightarrow i+1}^{(m)} = \alpha_{i}^{(m)}\left(1-\frac{v_{i}}{v_{\max}}\right)^{\eta^{(m)}} \left[1-\exp\left(-\frac{\max(0,\rho_{\max}-\Omega_{x,l})}{R_{c}}\right)\right]
\end{equation}

To prevent vehicles from driving into a completely jammed road, deceleration is controlled by a strong \textbf{Singular Asymptotic Barrier} $\mathcal{B}(\Omega_{x,l})$. A tiny number $\epsilon \approx 10^{-8}$ is used to prevent the computer from dividing by zero:
\begin{equation}
\mathcal{B}(\Omega_{x,l}) = \left(\frac{\rho_{\max}}{\max(\epsilon,\rho_{\max}-\Omega_{x,l})}\right)^{\eta^{(m)}}
\end{equation}
\begin{equation}
\lambda_{i\rightarrow i-1}^{(m)} = \left[ \omega_{0}^{(m)} + \sum_{n}\sum_{k<i}\beta_{i\rightarrow i-1}^{(m,n)}(v_{i}-v_{k}) \frac{w^{(n)} f_{k,x,l}^{(n)}}{\rho_{\max}} \right] \cdot \mathcal{B}(\Omega_{x,l})
\end{equation}

\subsection{Smooth Lane Changing ($\gamma$)}
To keep the computer calculations stable and smooth without sudden jumps, the lane-changing rate uses a smooth mathematical curve (Softplus). It is based on the density difference $y=\frac{\Omega_{x,l}-\Omega_{x,l'}}{\rho_{\max}}$ between the current lane and the target lane:
\begin{equation}
\gamma_{l\rightarrow l',i}^{(m)} = \frac{\kappa^{(m)}}{\omega}\ln[1+\exp(\omega\cdot y)] \cdot \left[1-\exp\left(-\frac{\max(0,\rho_{\max}-\Omega_{x,l'})}{R_{gap}}\right)\right]
\end{equation}

\section{Microscopic Reactions: Capture and Release}
\subsection{Capture Rate: Becoming Trapped ($Bf \rightarrow Bs$)}
A fast car becomes trapped only if it is blocked by trucks ahead while all adjacent escape lanes are full. First, we calculate the blocking effect of the trucks. The indicator function $\mathbf{1}_{v_i \ge v_k}$ guarantees that a car can only be captured by trucks moving slower than or equal to its own speed:
\begin{equation}
\mathcal{A}_{x,l}^{(i)} = \sum_{k \le i_{thr}} \mathbf{1}_{v_i \ge v_k} \left[ \omega_0^{(B,A)} + \beta^{(B,A)}(v_i - v_k) \right] \frac{w^{(A)} f_{k,x,l}^{(A)}}{\rho_{\max}}
\end{equation}

Next, we calculate the chance that the adjacent lane $l'$ is blocked. We use a strictly bounded curve ($\eta_{lat} > 0$) to ensure the value stays between 0 and 1. We also prevent cars from moving off the edge of the road ($G_{1 \to 0} \equiv 1, G_{L \to L+1} \equiv 1$):
\begin{equation}
G_{x,l\rightarrow l'} = \left( \frac{\min[\max(0, \Omega_{x,l'}), \rho_{\max}]}{\rho_{\max}} \right)^{\eta_{lat}}
\end{equation}

The Entrapment Factor $E_{x,l} = \prod_{l' \in \{l-1, l+1\}} G_{x,l\rightarrow l'}$ checks both adjacent lanes. A car is only captured if \textit{all} escape routes are blocked. The final capture rate strictly requires local blockage before it can be amplified by adjacent trucks:
\begin{equation}
\sigma_{x,l}^{(i)} = \sigma_0^{(B)} \cdot E_{x,l} \cdot \mathcal{A}_{x,l}^{(i)} \cdot \left[ 1 + \xi \sum_{l' \in \{l-1, l+1\}} \mathcal{A}_{x,l'}^{(i)} \right] \cdot \mathcal{B}(\Omega_{x,l})
\end{equation}

\subsection{Release Rate: Getting Free ($Bs \rightarrow Bf$)}
A trapped car becomes free again if the truck cluster disperses or an adjacent lane opens up. Let $\mathcal{S}_{x,l} = \sum_{k \le i_{thr}} \frac{w^{(A)} f_{k,x,l}^{(A)}}{\rho_{\max}}$ be the local density of trucks. To prevent the model from rapidly switching a car between trapped and free states, the effective blocking pressure $\tilde{\mathcal{S}}$ structurally mirrors the capture logic:
\begin{equation}
\tilde{\mathcal{S}}_{x,l} = E_{x,l} \cdot \mathcal{S}_{x,l} \cdot \left[ 1 + \xi \sum_{l' \in \{l-1, l+1\}} \mathcal{S}_{x,l'} \right]
\end{equation}
\begin{equation}
\mu_{x,l} = \mu_0^{(B)} \exp\left(-\frac{\tilde{\mathcal{S}}_{x,l}}{R_A}\right)
\end{equation}
\textit{Physical Axiom: If adjacent lanes become empty ($E \rightarrow 0$) or the trucks leave ($\mathcal{S} \rightarrow 0$), the blocking pressure disappears ($\tilde{\mathcal{S}} \rightarrow 0$). The release rate immediately jumps to its maximum $\mu_0^{(B)}$, and the car becomes free again.}

\section{Forward Movement and Flux Limits}
All vehicles move forward to the next cell at their actual instantaneous speed $v_i$. To prevent the computer from accidentally pushing too many cars into a full cell (which would crash the simulation), we use a global flux limiter $\alpha$. The continuous spatial demand to move forward is:
\begin{equation}
\Psi_{i,x\rightarrow x+1,l}^{(m)} = v_{i} f_{i,x,l}^{(m)} \cdot \left[1 - \exp\left(-\frac{\max(0, \rho_{\max} - \Omega_{x+1,l})}{R_{supply}}\right)\right]
\end{equation}

By defining the total equivalent spatial demand as $\mathcal{D}_{x\rightarrow x+1,l} = \frac{\Delta t}{\Delta x} \sum_{n, j} w^{(n)} \Psi_{j,x\rightarrow x+1,l}^{(n)}$, the limiter $\alpha$ restricts the flow if the demand exceeds the available space in the next cell:
\begin{equation}
\alpha_{x\rightarrow x+1,l} = 
\begin{cases} 
1, & \text{if } \mathcal{D}_{x\rightarrow x+1,l} = 0 \\
\min \left[ 1, \frac{\max(0, \rho_{\max} - \Omega_{x+1,l})}{\mathcal{D}_{x\rightarrow x+1,l}} \right], & \text{otherwise}
\end{cases}
\end{equation}
The final, safe amount of traffic moving forward is $\Phi_{i,x\rightarrow x+1,l}^{(m)} = \alpha_{x\rightarrow x+1,l} \cdot \Psi_{i,x\rightarrow x+1,l}^{(m)}$.

\section{Unified Governing PDEs}
To describe the traffic flow as a continuous fluid, we write the system as Partial Differential Equations (PDEs). The change in density over time ($\frac{\partial f}{\partial t}$) is balanced by the spatial traffic flow gradient ($\frac{\partial \Phi}{\partial x}$) and the internal state changes. We fully expand the equations to show exactly how vehicles change speeds ($\lambda$) and lanes ($\gamma$). Because the capture and release reactions happen at the exact same speed index $i$, no vehicles are accidentally created or destroyed.

For Class A (Trucks), the PDE includes spatial movement, speed changes, and lane changes:
\begin{equation}
\begin{aligned}
\frac{\partial f_{i,x,l}^{(A)}}{\partial t} + \frac{\partial \Phi_{i,x,l}^{(A)}}{\partial x} = 
& \sum_{j\in\{i-1,i+1\}} \left( \lambda_{j\rightarrow i}^{(A)} f_{j,x,l}^{(A)} - \lambda_{i\rightarrow j}^{(A)} f_{i,x,l}^{(A)} \right) \\
& + \sum_{l'\in\{l-1,l+1\}} \left( \gamma_{l'\rightarrow l, i}^{(A)} f_{i,x,l'}^{(A)} - \gamma_{l\rightarrow l', i}^{(A)} f_{i,x,l}^{(A)} \right)
\end{aligned}
\end{equation}

For Class Bf (Free-flowing Cars), the PDE adds the capture and release reactions. Cars gain density when released ($+\mu$) and lose density when captured ($-\sigma$):
\begin{equation}
\begin{aligned}
\frac{\partial f_{i,x,l}^{(Bf)}}{\partial t} + \frac{\partial \Phi_{i,x,l}^{(Bf)}}{\partial x} = 
& \sum_{j\in\{i-1,i+1\}} \left( \lambda_{j\rightarrow i}^{(Bf)} f_{j,x,l}^{(Bf)} - \lambda_{i\rightarrow j}^{(Bf)} f_{i,x,l}^{(Bf)} \right) \\
& + \sum_{l'\in\{l-1,l+1\}} \left( \gamma_{l'\rightarrow l, i}^{(Bf)} f_{i,x,l'}^{(Bf)} - \gamma_{l\rightarrow l', i}^{(Bf)} f_{i,x,l}^{(Bf)} \right) \\
& + \mu_{x,l} f_{i,x,l}^{(Bs)} - \sigma_{x,l}^{(i)} f_{i,x,l}^{(Bf)}
\end{aligned}
\end{equation}

For Class Bs (Trapped Cars), lane changing is strictly disabled. The PDE only includes spatial movement, constrained speed changes, and the capture/release reactions:
\begin{equation}
\begin{aligned}
\frac{\partial f_{i,x,l}^{(Bs)}}{\partial t} + \frac{\partial \Phi_{i,x,l}^{(Bs)}}{\partial x} = 
& \sum_{j\in\{i-1,i+1\}} \left( \lambda_{j\rightarrow i}^{(Bs)} f_{j,x,l}^{(Bs)} - \lambda_{i\rightarrow j}^{(Bs)} f_{i,x,l}^{(Bs)} \right) \\
& + \sigma_{x,l}^{(i)} f_{i,x,l}^{(Bf)} - \mu_{x,l} f_{i,x,l}^{(Bs)}
\end{aligned}
\end{equation}

\section{Mathematical Proof of Global Mass Conservation}
To guarantee the physical integrity of the model, we must rigorously prove that the introduction of the moving bottleneck subclasses (Bf and Bs) and their non-linear governing reactions do not violate the conservation of vehicles.

\begin{theorem}[Global Mass Conservation for Class B]
The total mass of Class B vehicles, defined as the spatial integration of both subclasses across all speed levels and lanes, is strictly conserved internally. Its temporal derivative is entirely determined by the net boundary fluxes.
\end{theorem}

\begin{proof}
Let the total mass of Class B within the network be $P^{(B)}$:
\begin{equation}
P^{(B)} = \sum_{x=1}^{X}\sum_{l=1}^{L}\sum_{i=1}^{N} \left( f_{i,x,l}^{(Bf)} + f_{i,x,l}^{(Bs)} \right) \Delta x
\end{equation}

Taking the time derivative of $P^{(B)}$ and substituting the governing partial differential equations yields:
\begin{equation}
\frac{d P^{(B)}}{dt} = \sum_{x=1}^{X}\sum_{l=1}^{L}\sum_{i=1}^{N} \left( \frac{\partial f_{i,x,l}^{(Bf)}}{\partial t} + \frac{\partial f_{i,x,l}^{(Bs)}}{\partial t} \right) \Delta x
\end{equation}

We expand this by evaluating the four fundamental operators separately:

\textbf{1. Capture and Release Reactions ($\sigma, \mu$):}
The reaction terms occur within the exact same speed index $i$. Summing the reaction terms for Bf and Bs yields:
\begin{equation}
\sum_{i=1}^{N} \left[ \left( \mu_{x,l} f_{i,x,l}^{(Bs)} - \sigma_{x,l}^{(i)} f_{i,x,l}^{(Bf)} \right) + \left( \sigma_{x,l}^{(i)} f_{i,x,l}^{(Bf)} - \mu_{x,l} f_{i,x,l}^{(Bs)} \right) \right] = \sum_{i=1}^{N} 0 = 0
\end{equation}
This perfect algebraic cancellation proves that the transition between free and trapped states is a strictly zero-sum exchange.

\textbf{2. Internal Speed Changes ($\lambda$):}
Because speed transitions are tridiagonal and sealed by strict limits ($\lambda_{1\rightarrow 0} \equiv 0$ and $\lambda_{N\rightarrow N+1} \equiv 0$), the summation over all discrete speed indices perfectly cancels:
\begin{equation}
\sum_{i=1}^{N} \sum_{j\in\{i-1,i+1\}} \left( \lambda_{j\rightarrow i} f_{j} - \lambda_{i\rightarrow j} f_{i} \right) = 0
\end{equation}

\textbf{3. Lateral Lane-Changing ($\gamma$):}
Similarly, symmetric lane transitions across all lanes $L$ cancel entirely due to the topological road boundaries ($G_{1 \rightarrow 0} \equiv 1$ and $G_{L \rightarrow L+1} \equiv 1$):
\begin{equation}
\sum_{l=1}^{L} \sum_{l'\in\{l-1,l+1\}} \left( \gamma_{l'\rightarrow l} f_{l'} - \gamma_{l\rightarrow l'} f_{l} \right) = 0
\end{equation}

\textbf{4. Spatial Advection ($\Phi$):}
The spatial difference forms a telescoping sum across the longitudinal domain $X$:
\begin{equation}
\sum_{x=1}^{X} \left( \Phi_{i,x-1\rightarrow x} - \Phi_{i,x\rightarrow x+1} \right) = \Phi_{i,0\rightarrow 1} - \Phi_{i,X\rightarrow X+1}
\end{equation}

Therefore, all internal transition operators identically evaluate to zero. The time derivative of the total mass reduces exclusively to the net flow across the network's outer boundaries:
\begin{equation}
\frac{d P^{(B)}}{dt} = \sum_{l=1}^{L}\sum_{i=1}^{N} \left( \Phi_{i,0\rightarrow 1}^{(Bf)} + \Phi_{i,0\rightarrow 1}^{(Bs)} - \Phi_{i,X\rightarrow X+1}^{(Bf)} - \Phi_{i,X\rightarrow X+1}^{(Bs)} \right)
\end{equation}
This strictly guarantees that no vehicles are unphysically created or destroyed within the dynamic multidimensional system.
\end{proof}

\section{Operator Splitting: The Calculation Sequence}
To prevent "ghost penetration" (where trapped high-speed vehicles unphysically move past the slow truck before the computer can slow them down), we calculate the system in a strict, step-by-step sequence during each time step $\Delta t$:

\begin{table}[h]
\centering
\begin{tabular}{llcl}
\toprule
\textbf{Phase} & \textbf{Operator} & \textbf{Stiffness} & \textbf{Numerical Solver} \\
\midrule
1 & Capture \& Release ($\sigma, \mu$) & \textbf{Extremely High} & Exact Analytical Matrix Exponential \\
2 & Internal Speed Changes ($\lambda$) & \textbf{Extremely High}  & Dynamic Algebraic Projection + Thomas \\
3 & Lateral Lane-Changing ($\gamma$) & Medium & Explicit Euler Integration \\
4 & Spatial Advection ($\Phi$) & Low & Explicit Euler + Godunov Limiter \\
\bottomrule
\end{tabular}
\end{table}

Let intermediate states be denoted by $(*, **, ***)$.

\textbf{Phase 1: Exact Capture and Release ($t \rightarrow t^*$)} \\
During this phase, cars do not move forward or change lanes, so the total density $\Omega_{x,l}$ is constant. This means the capture and release rates ($S^{(i)} = \sigma_{x,l}^{(i)} + \mu_{x,l}$) are also constant. To prevent the computer from failing if $S^{(i)} \to 0$, we use a safe mathematical function $\phi(z) = (1 - e^{-z})/z$. Let $F_{i,x,l} = f_{i,x,l}^{(Bf)}(t) + f_{i,x,l}^{(Bs)}(t)$ be the total cars at speed $i$. We can calculate the exact number of free and trapped cars instantly:
\begin{equation}
f_{i,x,l}^{(Bf), *} = f_{i,x,l}^{(Bf)}(t) \exp(-S^{(i)} \Delta t) + \mu_{x,l} F_{i,x,l} \Delta t \cdot \phi(S^{(i)} \Delta t)
\end{equation}
\begin{equation}
f_{i,x,l}^{(Bs), *} = F_{i,x,l} - f_{i,x,l}^{(Bf), *}
\end{equation}

\textbf{Phase 2: Instant Slowdown for Trapped Cars ($t^* \rightarrow t^{**}$)} \\
If a fast car is captured in Phase 1, it must slow down immediately before it is allowed to move forward. Let $\kappa_{x,l}^* \le i_{thr}$ be the actual speed of the trucks currently in the cell. We instantly shift all trapped cars that are going too fast down to the truck's speed:
\begin{equation}
f_{\kappa_{x,l}^*,x,l}^{(Bs), *} \leftarrow f_{\kappa_{x,l}^*,x,l}^{(Bs), *} + \sum_{j > \kappa_{x,l}^*} f_{j,x,l}^{(Bs), *} ; \quad f_{j,x,l}^{(Bs), *} \leftarrow 0 \quad (\forall j > \kappa_{x,l}^*)
\end{equation}
\textit{Why this is important:} This step completely empties the high-speed categories for trapped cars. This not only makes the computer's job much easier, but it also physically guarantees that trapped cars cannot unrealistically "jump" past the trucks in Phase 4. Because Phase 1 (Capture) and Phase 2 (Slowdown) happen \textit{before} Phase 4 (Forward Movement), trapped cars are safely slowed down before their positions are updated on the map. This perfectly simulates the moving bottleneck effect.

\end{document}